README

### 2022-09-28 rmd/supporting_info.Rmd
Update supporting_info.Rmd

We employed a multispecies Ricker model [@fowlerSpeciesDynamicsAlter2012]. In the basic formula without intentional release, the population density of species $i$ at time $t+1$, $N_{i,t+1}$, is modeled as:

@@

Under each simulation scenario, we ran `r attr(list_param[[1]], "n_sim")` time steps of community dynamics for $100$ values of release level $R$ (0 to 500 with an equal interval). We introduced 50 individuals of each species and allowed them to grow with no intentional release for the first `r unique(list_sim[[1]]$n_warmup)` time steps (initialization). After the initialization period, we released $R$ individuals of the enhanced species every time step over the next `r unique(list_sim[[1]]$n_burnin)` time steps to reach new equilibrium with intentional release (burn-in period). We continued the simulation run with intentional release and saved the last `r unique(list_sim[[1]]$n_timestep)` time steps, which were used to obtain the following summary statistics of the whole community $\sum_i^S N_i$: the coefficient of variation (CV), the temporal mean ($\mu$), and the SD ($\sigma$).

@@

As with the two-species model, we ran `r attr(list_param[[2]], "n_sim")` time steps of community dynamics with `r max(list_sim[[2]]$n_rep)` values of release level $R$ ($0$ to $500$ with a constant interval) for each simulation scenario. In this simulation, we initialized the community by introducing five individuals of each species and repeated this seeding every 10 time steps to account for possible complex dynamics of a ten-species community. The rest of the simulation procedure is identical to the two-species model. We summarized values of simulation parameters in **Table S**\@ref(tab:tab-para).

@@

**Time-series data.** We assembled time-series fish data at `r n_distinct(d0$site_id)` sites within `r n_distinct(d0$river)` protected watersheds of Hokkaido Island, Japan. The Hokkaido Research Organization leads a long-term monitoring program at these watersheds, and the data are published as annual reports [@salmonandfreshwaterfisheriesresearchinstitutehokkaidoresearchorganizationAnnualReportMonitoring2021]. The program began in 1963, but an effective, standardized sampling method has been implemented since 1999 (two-pass sampling with a combination of electrofishing and cast net). Most data were collected in summer with irregular interannual intervals (1- to 3-year intervals for most cases), and sampling efforts were quantified by sampling area (average: `r op(mean(df_area$area))` $\pm$ `r op(sd(df_area$area))` m^2^). We confined our analysis to the sites that meet the following criteria: (i) the observation span (from the first to the last year of observation) exceeds 10 years, (ii) the number of observation years exceeds five years, and (iii) masu salmon is observed at least twice during the observation period. As a result, we used time-series data at `r n_distinct(df_fish$site_id)` sites within `r n_distinct(df_fish$river)` watersheds from 1999 to 2019. Summed abundance of first and second passes was used in the following analysis. **Table S**\@ref(tab:tab-splist) summarizes observed species in these watersheds.

@@

```{=tex}
\begin{align}
\begin{split}
&\ln n_{s,t} = \xi_{1,s} + \xi_{2,s} \ln n'_{s,t-1} + \varepsilon_{s,t}^{\text{state}}\\
&\ln n'_{s,t-1} = \frac{\sum_{q=1}^Q (\tau^{q} \ln n_{s,t-q})}{\sum_{q=1}^Q \tau^{q}}
\end{split}
(\#eq:state)
\end{align}
```
where $\xi_{1,s}$ is the site-specific constant at site $s$, $\xi_{2,s}$ the parameter characterizing an autoregressive (AR) process, and $\varepsilon_{s,t}^{\text{state}}$ the process error that follows a normal distribution as $\varepsilon_{s,t}^{\text{state}} \sim \text{Normal}(0, \sigma_{\text{state},s}^2)$. The parameters $\xi_{1,s}$ and $\xi_{2,s}$ were drawn from a multi-variate normal distribution as $\pmb{\xi} \sim \text{MVN}(\pmb{\theta_{\xi}}, \pmb{\Omega_{\xi}})$, where $\pmb{\xi}$ is the matrix of $\xi_{1,s}$ and $\xi_{2,s}$. Equation \@ref(eq:state) is structurally equivalent to a $Q$-order AR model; however, we expressed the part of AR parameters as a geometric-series $\tau^q$ to reduce the number of parameters (our model has two AR parameters $\xi_{2,s}$ and $\tau$ regardless of order $Q$). The parameter $\tau$ controls how fast the density influence attenuates towards the past [@takagiTimescaleDependencyHost2015]. In this study, we set $Q=3$ given the life span of the study species (typically 1 -- 3 years). We used median estimates of fish density $n_{s,t}$ to calculate the temporal CV, mean ($\mu$), and SD ($\sigma$) for each site. For a whole community, we summed the densities of enhanced and unenhanced species. We summarized the reconstructed community dynamics in **Figures S**\@ref(fig:fig-d1) -- **S**\@ref(fig:fig-d3).

@@

We used linear regression to quantify the impact of intentional release on community dynamics. Although our focus is intentional release, each model included climatic and local abiotic variables to account for important environmental differences among sites. Specifically, we developed the following linear regression model taking either species richness (the number of species present during the observation period), mean, or SD as a response variable $y_s$.

```{=tex}
\begin{align}
\left\{
\begin{aligned}
y_s &\sim \text{Poisson}(\lambda_{y,s} \exp(\varepsilon_{\lambda,s})) &&\text{for species richness}\\
\ln \pmb{y_s} &\sim \text{MVN}(\pmb{\theta_{y,s}}, \pmb{\Omega_y}) &&\text{for mean and SD}
\end{aligned}
\right.
 (\#eq:reg)
\end{align}
```
In the poisson model for species richness, $\exp(\varepsilon_{\lambda,s})$ is the error term accounting for overdispersion ($\varepsilon_{\lambda,s} \sim \text{Normal}(0, \sigma_{\lambda}^2)$). In the multi-variate normal model, $\pmb{y_s}$ represents a vector of mean and SD at site $s$, which is modeled as a random variable drawn from a multi-variate normal distribution with the variance-covariance matrix $\pmb{\Omega_y}$. The expected means $\lambda_{y,s}$ and $\pmb{\theta_s}$ were related to linear predictors as follows:

@@

```{=tex}
\begin{align}
\begin{split}
R'_s &= \eta_{w,s'} R_w\\
\sum_{s'}^{S_w + 1} \eta_{w,s'} &= 1
(\#eq:effr)
\end{split}
\end{align}
```
In Equation \@ref(eq:effr), $\eta_{w,s'}$ represents the effective proportion of total release affecting the community at site $s'$ within watershed $w$. The summation of $\eta_{w,s'}$ over $S_w+1$ is assumed to be a unity because non-negligible portion of release may escape into unsurveyed areas. Since the values of $\eta_{w,s'}$ are not available in our dataset, we estimated them through a stochastic search while fitting the regression model to the data within a Bayesian framework (see **Model fitting**). Other site-level predictors include upstream watershed area (log-transformed), air temperature, precipitation, and forest land use. Urban and agricultural land use were omitted because of either a limited value range (**Figure S**\@ref(fig:fig-env)) or a strong correlation with forest land use (**Figure S**\@ref(fig:fig-cor)). The watershed-specific intercept was related to watershed-level predictors as:

@@

Since the CV is a additive function of mean $\mu$ and SD $\sigma$ in a log scale, we can derive the effects of environmental factors on the CV from Equation \@ref(eq:model):

@@

```{=tex}
\begin{align}
\varepsilon_{i,t}^{\text{state}}=\sum_{d}^{D}\zeta_{t,d}\delta_{d,i} + I_{ij}\epsilon_{i,t}
\end{align}
```
where $D$ is the number of latent factors, $\epsilon_{i,t} \sim \text{Normal}(0, \sigma_{\epsilon,i}^2)$, and $I_{ij}$ is the identity matrix ($I_{ii}=1$ and $I_{ij, i \ne j} = 0$). With this formulation, the matrix $\pmb{\Omega_{\varepsilon}}$ can be described as $\pmb{\Omega_{\varepsilon}}=\pmb{\Delta}^T\pmb{\Delta} + \text{diag}(\sigma_{\epsilon,i}^2)$, where $\pmb{\Delta}$ is the matrix of factor loadings $\delta_{d,i}$. This model structure has a merit of reduced number of parameters while capturing the correlated nature of species responses to environmental fluctuations [@ovaskainenHowAreSpecies2017]. We set $D=2$ in our model. We reported the estimates of $\alpha_{ij}$ so the values are comparable to those used in the theoretical analysis.

@@

```{r fig-2sp, fig.width=11, fig.height=8, fig.cap = paste0("Theoretical predictions in a 2-species community. **(A)** Spearman's rank correlations between intentional release and community dynamics in a strong competition scenario ($\\bar{\\alpha}=$ ", max(list_param[[1]]$alpha) ,"). Rows distinguish different summary statistics of community dynamics (top: CV, middle: mean, bottom: SD) while columns distinguish deterministic ($\\sigma_{\\varepsilon}=$ ", min(list_param[[1]]$sd_env) ,") and stochastic simulations ($\\sigma_{\\varepsilon}=$ ", max(list_param[[1]]$sd_env) ,"). Cells indicate 400 combinations of carrying capacity and intrinsic growth rate (20 values each). Colors are proportional to Spearman's rank correlations. Dots indicate parameter combinations that are used in the whole community simulation. **(B)** Examples of community dynamics with and without intentional release. Community dynamics diverge at the broken vertical lines, the time step at which intentional release begins. Rows distinguish intrinsic growth rates while columns distinguish deterministic and stochastic simulations. Colors indicate release levels, and line types differ in carrying capacities.")}

@@

  cap[i] <- paste0("Theoretical predictions for the release effect on community dynamics. Rows represent different response variables, and columns show distinct simulation scenarios with different strength of interspecific competition ($\\bar{\\alpha}$) and relative fitness of captive-bred individuals ($f_R$). Other parameters are: ",
                       "number of species $S$ = ", df_param$n_species[i], "; intrinsic growth rate of an enhanced species $r_1$ = ", df_param$r1[i], "; maximum intrinsic growth rate of unenhanced species $r_{max}$ = ", df_param$r_max[i], "; environmental stochasticity $\\sigma_{\\varepsilon}$ = ", df_param$sd_env[i], "; carrying capacity $K$ = ", df_param$k[i], ".")

>>>>

We employed a multi-species Ricker model [@fowlerSpeciesDynamicsAlter2012]. In the basic formula without intentional release, the population density of species $i$ at time $t+1$, $N_{i,t+1}$, is modeled as:

@@

Under each simulation scenario, we ran `r attr(list_param[[1]], "n_sim")` timesteps of community dynamics for $100$ values of release level $R$ (0 to 500 with an equal interval). We introduced 50 individuals of each species and allowed them to grow with no intentional release for the first `r unique(list_sim[[1]]$n_warmup)` timesteps (initialization). After the initialization period, we released $R$ individuals of the enhanced species every timestep over the next `r unique(list_sim[[1]]$n_burnin)` timesteps to reach a new equilibrium with the intentional release (burn-in period). We continued the simulation run with intentional release and saved the last `r unique(list_sim[[1]]$n_timestep)` timesteps, which were used to obtain the following summary statistics of the whole community $\sum_i^S N_i$: the coefficient of variation (CV), the temporal mean ($\mu$), and the SD ($\sigma$).

@@

As with the two-species model, we ran `r attr(list_param[[2]], "n_sim")` timesteps of community dynamics with `r max(list_sim[[2]]$n_rep)` values of release level $R$ ($0$ to $500$ with a constant interval) for each simulation scenario. In this simulation, we initialized the community by introducing five individuals of each species and repeated this seeding every ten timesteps to account for possible complex dynamics of a ten-species community. The rest of the simulation procedure is identical to the two-species model. We summarized values of simulation parameters in **Table S**\@ref(tab:tab-para).

@@

**Time-series data.** We assembled time-series fish data at `r n_distinct(d0$site_id)` sites within `r n_distinct(d0$river)` protected watersheds of Hokkaido Island, Japan. The Hokkaido Research Organization leads a long-term monitoring program at these watersheds, and the data are published as annual reports [@salmonandfreshwaterfisheriesresearchinstitutehokkaidoresearchorganizationAnnualReportMonitoring2021]. The program began in 1963, but an effective, standardized sampling method has been implemented since 1999 (two-pass sampling with a combination of electrofishing and cast net). Most data were collected in summer with irregular interannual intervals (1- to 3-year intervals for most cases), and sampling efforts were quantified by sampling area (average: `r op(mean(df_area$area))` $\pm$ `r op(sd(df_area$area))` m^2^). We confined our analysis to the sites that meet the following criteria: (i) the observation span (from the first to the last year of observation) exceeds ten years, (ii) the number of observation years exceeds five years, and (iii) masu salmon is observed at least twice during the observation period. As a result, we used time-series data at `r n_distinct(df_fish$site_id)` sites within `r n_distinct(df_fish$river)` watersheds from 1999 to 2019. The summed abundance of first and second passes was used in the following analysis. **Table S**\@ref(tab:tab-splist) summarizes observed species in these watersheds.

@@

```{=tex}
\begin{align}
\begin{split}
&\ln n_{s,t} = \xi_{1,s} + \xi_{2,s} \ln n'_{s,t-1} + \varepsilon_{s,t}^{\text{state}}\\
&\ln n'_{s,t-1} = \frac{\sum_{q=1}^Q (\tau^{q} \ln n_{s,t-q})}{\sum_{q=1}^Q \tau^{q}}
\end{split}
(\#eq:state)
\end{align}
```
where $\xi_{1,s}$ is the site-specific constant at site $s$, $\xi_{2,s}$ the parameter characterizing an autoregressive (AR) process, and $\varepsilon_{s,t}^{\text{state}}$ the process error that follows a normal distribution as $\varepsilon_{s,t}^{\text{state}} \sim \text{Normal}(0, \sigma_{\text{state},s}^2)$. The parameters $\xi_{1,s}$ and $\xi_{2,s}$ were drawn from a multivariate normal distribution as $\pmb{\xi} \sim \text{MVN}(\pmb{\theta_{\xi}}, \pmb{\Omega_{\xi}})$, where $\pmb{\xi}$ is the matrix of $\xi_{1,s}$ and $\xi_{2,s}$. Equation \@ref(eq:state) is structurally equivalent to a $Q$-order AR model; however, we expressed the part of AR parameters as a geometric-series $\tau^q$ to reduce the number of parameters (our model has two AR parameters $\xi_{2,s}$ and $\tau$ regardless of order $Q$). The parameter $\tau$ controls how fast the density influence attenuates towards the past [@takagiTimescaleDependencyHost2015]. In this study, we set $Q=3$ given the life span of the study species (typically 1 -- 3 years). We used median estimates of fish density $n_{s,t}$ to calculate the temporal CV, mean ($\mu$), and SD ($\sigma$) for each site. For a whole community, we summed the densities of enhanced and unenhanced species. We summarized the reconstructed community dynamics in **Figures S**\@ref(fig:fig-d1) -- **S**\@ref(fig:fig-d3).

@@

We used linear regression to quantify the impact of the intentional release on community dynamics. Although our focus is intentional release, each model included climatic and local abiotic variables to account for important environmental differences among sites. Specifically, we developed the following linear regression model taking either species richness (the number of species present during the observation period), mean, or SD as a response variable $y_s$.

```{=tex}
\begin{align}
\left\{
\begin{aligned}
y_s &\sim \text{Poisson}(\lambda_{y,s} \exp(\varepsilon_{\lambda,s})) &&\text{for species richness}\\
\ln \pmb{y_s} &\sim \text{MVN}(\pmb{\theta_{y,s}}, \pmb{\Omega_y}) &&\text{for mean and SD}
\end{aligned}
\right.
 (\#eq:reg)
\end{align}
```
In the poisson model for species richness, $\exp(\varepsilon_{\lambda,s})$ is the error term accounting for overdispersion ($\varepsilon_{\lambda,s} \sim \text{Normal}(0, \sigma_{\lambda}^2)$). In the multivariate normal model, $\pmb{y_s}$ represents a vector of mean and SD at site $s$, which is modeled as a random variable drawn from a multivariate normal distribution with the variance-covariance matrix $\pmb{\Omega_y}$. The expected means $\lambda_{y,s}$ and $\pmb{\theta_s}$ were related to linear predictors as follows:

@@

```{=tex}
\begin{align}
\begin{split}
R'_s &= \eta_{w,s'} R_w\\
\sum_{s'}^{S_w + 1} \eta_{w,s'} &= 1
(\#eq:effr)
\end{split}
\end{align}
```
In Equation \@ref(eq:effr), $\eta_{w,s'}$ represents the effective proportion of total release affecting the community at site $s'$ within watershed $w$. The summation of $\eta_{w,s'}$ over $S_w+1$ is assumed to be unity because a non-negligible portion of release may escape into unsurveyed areas. Since the values of $\eta_{w,s'}$ are not available in our dataset, we estimated them through a stochastic search while fitting the regression model to the data within a Bayesian framework (see **model fitting**). Other site-level predictors include upstream watershed area (log-transformed), air temperature, precipitation, and forest land use. Urban and agricultural land use was omitted because of either a limited value range (**Figure S**\@ref(fig:fig-env)) or a strong correlation with forest land use (**Figure S**\@ref(fig:fig-cor)). The watershed-specific intercept was related to watershed-level predictors as:

@@

Since the CV is an additive function of mean $\mu$ and SD $\sigma$ in a log scale, we can derive the effects of environmental factors on the CV from Equation \@ref(eq: model):

@@

```{=tex}
\begin{align}
\varepsilon_{i,t}^{\text{state}}=\sum_{d}^{D}\zeta_{t,d}\delta_{d,i} + I_{ij}\epsilon_{i,t}
\end{align}
```
where $D$ is the number of latent factors, $\epsilon_{i,t} \sim \text{Normal}(0, \sigma_{\epsilon,i}^2)$, and $I_{ij}$ is the identity matrix ($I_{ii}=1$ and $I_{ij, i \ne j} = 0$). With this formulation, the matrix $\pmb{\Omega_{\varepsilon}}$ can be described as $\pmb{\Omega_{\varepsilon}}=\pmb{\Delta}^T\pmb{\Delta} + \text{diag}(\sigma_{\epsilon,i}^2)$, where $\pmb{\Delta}$ is the matrix of factor loadings $\delta_{d,i}$. This model structure has the merit of the reduced number of parameters while capturing the correlated nature of species responses to environmental fluctuations [@ovaskainenHowAreSpecies2017]. We set $D=2$ in our model. We reported the estimates of $\alpha_{ij}$, so the values are comparable to those used in the theoretical analysis.

@@

```{r fig-2sp, fig.width=11, fig.height=8, fig.cap = paste0("Theoretical predictions in a 2-species community. **(A)** Spearman's rank correlations between intentional release and community dynamics in a strong competition scenario ($\\bar{\\alpha}=$ ", max(list_param[[1]]$alpha) ,"). Rows distinguish different summary statistics of community dynamics (top: CV, middle: mean, bottom: SD) while columns distinguish deterministic ($\\sigma_{\\varepsilon}=$ ", min(list_param[[1]]$sd_env) ,") and stochastic simulations ($\\sigma_{\\varepsilon}=$ ", max(list_param[[1]]$sd_env) ,"). Cells indicate 400 combinations of carrying capacity and intrinsic growth rate (20 values each). Colors are proportional to Spearman's rank correlations. Dots indicate parameter combinations that are used in the whole community simulation. **(B)** Examples of community dynamics with and without intentional release. Community dynamics diverge at the broken vertical lines, the timestep at which intentional release begins. Rows distinguish intrinsic growth rates, while columns distinguish deterministic and stochastic simulations. Colors indicate release levels, and line types differ in carrying capacities.")}

@@

  cap[i] <- paste0("Theoretical predictions for the release effect on community dynamics. Rows represent different response variables, and columns show distinct simulation scenarios with different strengths of interspecific competition ($\\bar{\\alpha}$) and relative fitness of captive-bred individuals ($f_R$). Other parameters are: ",
                       "number of species $S$ = ", df_param$n_species[i], "; intrinsic growth rate of an enhanced species $r_1$ = ", df_param$r1[i], "; maximum intrinsic growth rate of unenhanced species $r_{max}$ = ", df_param$r_max[i], "; environmental stochasticity $\\sigma_{\\varepsilon}$ = ", df_param$sd_env[i], "; carrying capacity $K$ = ", df_param$k[i], ".")
